Virtualization of a Real-Time Operating System for Robot Control with a Focus on Real-Time Compliance

### 2024-09-01 Masterarbeit_Robotik_Pamuk.tex
Final Correction

%Formatieren des Quellcodeverzeichnisses
\makeatletter
% Setzen der Bezeichnungen für das Quellcodeverzeichnis/Abkürzungsverzeichnis in Abhängigkeit von der eingestellten Sprache
\providecommand\listacroname{}
\@ifclasswith{twbook}{english}
{%
    \renewcommand\lstlistingname{Code}
    \renewcommand\lstlistlistingname{List of Code}
    \renewcommand\listacroname{List of Abbreviations}
}{%
    \renewcommand\lstlistingname{Quellcode}
    \renewcommand\lstlistlistingname{Quellcodeverzeichnis}
    \renewcommand\listacroname{Abkürzungsverzeichnis}
}
% Wenn die Option listof=entryprefix gewählt wurde, Definition des Entyprefixes für das Quellcodeverzeichnis. Definition des Macros listoflolentryname analog zu listoflofentryname und listoflotentryname der KOMA-Klasse
\@ifclasswith{scrbook}{listof=entryprefix}
{%
    \newcommand\listoflolentryname\lstlistingname
}{%
}
\makeatother
\newcommand{\listofcode}{\phantomsection\lstlistoflistings}

@@

\bigskip \noindent Insgesamt bietet diese Arbeit eine umfassende Methodik, um ein virtualisiertes Gastsystem in einem Hostsystem mit deterministischem Verhalten echtzeitfähig zu machen.}
\schlagworte{Virtualisierung, Echtzeitsysteme, Latenzreduktion, Robotiksteuerung}

@@

\chapter{Introduction}\label{cha:introduction}
In today's industrial production and automation, robot systems are well established and of crucial importance. Robots must react to their environment and perform time-critical tasks within strict time constraints. Delays or errors can have catastrophic consequences in some cases. Modern operating systems nowadays are categorized into two types:  General-Purpose Operating Systems (GPOS) and Real-Time Operating Systems (RTOS). The key difference is whether the system is time-critical or not~\cite{canbazPerformanceAnalysisRealtime2022}. In a traditional GPOS, such as Windows or Linux-based distributions, a high-priority task cannot interrupt a kernel operation, as illustrated in Figure~\ref{fig:kernel_generic}. This makes them not suitable for real-time requirements, as they cannot guarantee deterministic execution times.

@@

\noindent In these real-time operating systems, task scheduling is based on so-called priority-based preemptive scheduling~\cite{buttazzoHardRealtimeComputing2024}. Each task in a software application is assigned a priority. A higher priority means that a faster response is required. Preemptive task scheduling ensures a very fast response. Preemptive means that the scheduler can stop a currently running task at any point if it recognizes that another task needs to be executed immediately. The basic rule on which priority-based preemptive scheduling is based is that the task with the highest priority that is ready to run is always the task that must be executed. So if both a task with a lower priority and a task with a higher priority are ready to run, the scheduler ensures that the task with the higher priority runs first. The lower priority task is only executed once the higher priority task has been processed.

@@

\begin{itemize}
	\item This work was written at SIGMATEK GmbH \& Co KG, a company that researches, develops, produces, and distributes automation technology for industrial machinery and plant engineering. SIGMATEK uses its own custom Linux-based operating system, to be run on their self-manufactured CPUs. This operating system will be referred to as ``Salamander 4'' in this work. The details of Salamander 4 are explained in Chapter~\ref{sec:guest_operating_system}.
	\item Salamander 4 was created with Yocto~\cite{WelcomeYoctoProject}, an open-source project that helps develop custom Linux-based systems. Yocto is discussed in Section~\ref{sec:yocto}.
	\item Salamander 4 is virtualized through Quick Emulator (QEMU), an open-source emulator and virtualizer that allows running different operating systems on a computer. QEMU is described in Section~\ref{sec:qemu}.
	\item Salamander 4 employs hard real-time with Xenomai 3~\cite{XenomaiXenomai}, a real-time framework for Linux that enables deterministic and low-latency performance. Xenomai 3 is detailed in Section~\ref{sec:xenomai}.
	\item The goal is to virtualize Salamander 4 and approach the performance of bare metal (hardware) through real-time performance tunings, focusing on latency requirements set by SIGMATEK. These modifications are the subject of Chapter~\ref{sec:real-time_tunings}.
	\item The \texttt{latency} tool of the Xenomai suite to measure the scheduling latency of the real-time OS was suggested to be used by SIGMATEK. The goal was set to achieve latency values below 50 microseconds in the duration of the measurement.
	\item SIGMATEK uses the VARAN bus for communication between machines and systems, a real-time Ethernet network that was specially developed for industrial automation. Its operating principle is outlined in Section~\ref{sec:varan}.
	\item For the purpose of replacing the physical CPU, SIGMATEK provided a PCI Insert Card module PCV 522, that was plugged into the PC, as explained in Subsection~\ref{subsec:setup_experiment_virtualized}. This component could then be used to communicate with the Input/Output peripherals, just like the physical CPU could when the I/O modules were attached to it. Subsequently, the PCV 522 module of SIGMATEK was connected with the Pulse Width Module PW 022 of SIGMATEK over the VARAN bus and the VI 021 module of SIGMATEK serving as the power supply.
	\item The robotic application from Chapter~\ref{sec:robotic_application} was written in Lasal Class 2, an object-oriented programming tool of SIGMATEK and part of SIGMATEK's Engineering Platform LASAL, with client-server technology and graphic representation.
\end{itemize}

@@

\bigskip \noindent \citeauthor{perneelRealtimeCapabilitiesStandard2015}~\cite{perneelRealtimeCapabilitiesStandard2015} mention in their work that Linux was initially developed as a general-purpose operating system without real-time applications in mind. They talk about the evolution of real-time behavior in the standard Linux kernel and state that it has recently gained popularity among the real-time community, largely due to its open-source nature and stability. The authors explain how to transform Linux into a real-time operating system by implementing the PREEMPT-RT patches. The paper also emphasizes some kernel configurations that ensure that the system's real-time behavior is preserved during runtime. Even though these modifications to the kernel enable soft real-time performance, the authors underline the fundamental rule in real-time software, that latency improvements have a negative impact on throughput and kernel performance. This is due to the added overhead of the \texttt{CONFIG\_PREEMPT\_RT} option. They concluded with their Linux build and a testing application that Linux can be a viable option for real-time applications, if it is correctly configured. \citeauthor{reghenzaniRealTimeLinuxKernel2020}~\cite{reghenzaniRealTimeLinuxKernel2020} provide an extensive overview of the real-time Linux kernel research, specifically on the PREEMPT RT patch. \citeauthor{adamRealTimePerformanceResponse2021}~\cite{adamRealTimePerformanceResponse2021} applied the PREEMPT\_RT patch on both Raspberry Pi 3 and BeagleBone Black and achieved significantly lower latencies compared to standard Linux kernels.

@@

\bigskip \noindent In the course of writing this thesis, the white paper "Real-Time Performance Tuning Best Practice Guidelines for KVM-Based Virtual Machines"~\cite{RealTimePerformanceTuning2022} by Intel has been instrumental. Given that the working computer used for this work is equipped with an Intel processor, the guidelines presented in this document were particularly relevant and valuable for optimizing real-time performance in KVM-based virtual machines. Chapter~\ref{sec:real-time_tunings} contains an extensive tuning process, which involves configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS itself. The document concludes that the described tuning methods effectively improve real-time performance for KVM-based VMs, even under heavy workloads.

@@

\bigskip \noindent \citeauthor{mckenneyRealTimeVs2008}~\cite{mckenneyRealTimeVs2008} addresses the overheads associated with real-time Linux when using KVM and QEMU. The author lists several sources of overhead, including memory locking to avoid page-fault latencies, increased overhead of locking and interrupts due to more-aggressive preemption, threaded interrupts that permit long-running interrupt handlers to be preempted by high-priority real-time processes, real-time task scheduling that requires global scheduling, high-resolution timers for tens-of-microseconds accuracy and precision, and preemptible RCU which has slightly higher read-side overhead than Classic RCU. This information helped understand the trade-offs involved in real-time Linux systems.

\bigskip \noindent \citeauthor{deoliveiraDemystifyingRealTimeLinux2020}~\cite{deoliveiraDemystifyingRealTimeLinux2020} critique the traditional \texttt{cyclictest} tool used for measuring scheduling latency, highlighting its limitations due to its black-box approach and lack of theoretical grounding. This contributed to the decision not to use this tool to measure the scheduling latency in this work. Instead, the \texttt{latency} tool was used, as was determined in Chapter~\ref{sec:application_context}.

@@

\bigskip \noindent Virtualization promises to overcome these challenges~\cite{queirozContainerbasedVirtualizationRealtime2024}~\cite{bhardwajVirtualizationCloudComputing2021}~\cite{cinqueEvaluatingVirtualizationFog2023}. However, virtualization also has its own challenges and pitfalls. The additional layers of abstraction can lead to increased overhead and latency~\cite{casiniLatencyAnalysisVirtualization2021}~\cite{zhangEvaluatingOptimizingVirtualization2010}. This is not desired and can be particularly problematic for real-time systems that require strict timing constraints. Making sure that tasks run predictably and on time in a virtualized environment can be challenging. The hypervisor and other virtual components can cause variations in how long tasks take to execute~\cite{garcia-vallsChallengesRealtimeVirtualization2014}. When there are multiple virtual machines (VM) sharing the same physical resources, the VMs can affect each other's performance and consequently violate real-time constraints~\cite{queirozContainerbasedVirtualizationRealtime2024}.

@@

\bigskip \noindent In this master's thesis, the task is the virtualization of a real-time operating system to control a robot, with a focus on compliance with real-time determinism. Chapter~\ref{cha:methodology} explains in detail the procedure, how this was tackled. As already established in Section~\ref{sec:application_context}, Salamander 4 is built with Yocto, employs hard real-time with Xenomai 3, and is virtualized through QEMU/KVM. The initial latency values are measured with the \texttt{latency} program of the Xenomai tool suite. The initial situation is described in Section~\ref{sec:starting_situation}. In a nutshell, there is a significant initial gap in latency statistics between the virtualized Salamander 4 and the Salamander 4 on bare metal. For this reason, an extensive tuning process is carried out to achieve real-time performance and determinism. These modifications involve configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 guest OS itself. The individual configurations are discussed in detail in Chapter~\ref{sec:real-time_tunings} and the modifications are justified with a clear explanation.

\section{Objective}
The goal is to bring the latency of the virtualized Salamander 4 closer to that of the hardware, thereby providing a comprehensive blueprint for making a virtualized guest system in a host system real-time capable with deterministic behavior. Section~\ref{sec:starting_situation} outlines the initial situation of this master's thesis and serves as the foundation for the real-time performance tunings discussed in Section~\ref{sec:real-time_tunings}. The objective is to achieve latency values below 50 microseconds during the measurement period through real-time tunings, with no outliers whatsoever. To determine and validate the performance of the tunings, the \texttt{latency} tool is used, as stated in Section~\ref{sec:application_context}. On top of that, the aim is to use a robotic application in a practical scenario to demonstrate and validate that the tunings significantly reduce latency and improve determinism compared to the unmodified virtualization.

@@

This chapter describes in detail all the theoretical concepts and boundary conditions as well as practical methods that contributed to achieving the objectives of this master's thesis.

@@

\subsection{Structure}
\noindent Salamander 4 is the proprietary operating system of SIGMATEK. It is based on Linux version 5.15.94 and integrates Xenomai 3.2, a real-time development environment~\cite{XenomaiXenomai}. Salamander 4 is a 64-bit system, which refers to the x86\_64 architecture. The real-time behaviour is achieved through the use of Symmetric Multi-Processing (SMP) and Preemptive Scheduling (PREEMPT). In addition, it uses IRQPIPE to process interrupts in a way that meets the real-time requirements of the system. The output of the command~\texttt{uname -a} can be observed in code~\ref{output:uname_a_virt}.

\vspace{2em}
\begin{lstlisting}[name={System Information},label={output:uname_a_virt}]
root@sigmatek-core2:~# uname -a
Linux sigmatek-core2 5.15.94 #1 SMP PREEMPT IRQPIPE Tue Feb 14 18:18:05 UTC 2023 x86_64 GNU/Linux
\end{lstlisting}

@@

\begin{itemize}
	\item \textbf{Operating system}: The operating system in a LASAL CPU manages the hardware and software resources of the system. It is provided in a completely PC-compatible manner, working with a standard PC BIOS.
	\item \textbf{Loader}: The loader is a part of the operating system that is responsible for loading programs from executables into memory, preparing them for execution and then running them.
	\item \textbf{Hardware classes}: Hardware classes in LASAL represent the different types of hardware components that can be controlled by the LASAL CPU. They help organize and manage the hardware components in a modular manner. The graphical hardware editor in LASAL allows for a true-to-detail simulation of the actual hardware.
	\item \textbf{Applications}: Applications are developed using LASAL CLASS 2~\cite{EngineeringToolLASAL}, a solution for automation tasks that supports object-oriented programming and design in compliance with IEC 61131-3.
\end{itemize}

@@

QEMU~\cite{QEMU} is an open-source emulator and virtualizer that allows running operating systems and programs for one machine on a different machine. It can emulate different hardware architectures and is used in this thesis to emulate Salamander 4. When QEMU is used together with Kernel-based Virtual Machine (KVM), it can provide near-native performance by running code directly on the host CPU. The QEMU script used to emulate Salamander 4 and the meaning of the options are described in detail in Subsection \ref{subsec:salamander4-virtualization}.

@@

\vspace{2em}
\begin{lstlisting}[name={Contents of QEMU Folder for Salamander 4},label={lst:ls}]
    sigma_ibo@localhost:~/Desktop/salamander-image$ ls -1
    bzImage
    drive-c
    ovmf.code.qcow2
    qemu_def.sh
    salamander-image-sigmatek-core2.ext4
    stek-drive-c-image-sigmatek-core2.tar.gz
    vmlinux
    \end{lstlisting}

@@

\begin{itemize}
	\item \textbf{bzImage}: Compressed Linux kernel image that is loaded by QEMU at system start. The short form ``bz`` stands for big-zipped.
	\item \textbf{drive-c}: Directory serving as C drive for QEMU, created and filled by qemu\_def.sh script.
	\item \textbf{ovmf.code.qcow2}: Firmware file for QEMU that enables UEFI boot process. ~\texttt{OVMF} stands for Open Virtual Machine Firmware,~\texttt{qcow2} is a format for disk image files used by QEMU, it stands for ``QEMU Copy On Write version 2''.
	\item \textbf{qemu\_def.sh}: Shell script that starts QEMU with required parameters to boot Salamamder 4 OS. It is described later in more detail.
	\item \textbf{salamander-image-sigmatek-core2.ext4}: Disk image of the Salamander 4 OS for the Sigmatek Core 2 platform. It uses the~\texttt{ext4} file system and serves as the root file system in the QEMU virtual machine, acting as the virtual hard drive.
	\item \textbf{stek-drive-c-image-sigmatek-core2.tar.gz}: Compressed tarball containing a pre-configured environment for the Salamander 4 OS. It is unpacked and sets up the~\texttt{drive-c/} directory with system and log files in the~\texttt{qemu\_def.sh} script.
	\item \textbf{vmlinux}: Uncompressed Linux kernel image, typically used for debugging.
\end{itemize}

\noindent The initial QEMU script after the custom Yocto build and the starting point for this work is shown in Code~\ref{script:qemu_def}. This script is used to start QEMU with required parameters to boot Salamander 4 OS. It will be adjusted in Chapter~\ref{sec:real-time_tunings} in order to accompany real-time performance tunings.

\vspace{2em}
\begin{lstlisting}[name={QEMU script for starting Salamander 4 Virtualization},label={script:qemu_def}]
	#!/bin/sh

@@

\noindent  The second approach, called Mercury and shown in Figure~\ref{fig:mercury}, relies on the real-time capabilities already present in the native Linux kernel. Often, applications require the PREEMPT-RT~\cite{RealtimePreempt_rt_versionsWiki} patch to augment the mainline kernel's real-time responsiveness and minimize jitter, but this is not mandatory and depends on the application's specific requirements for responsiveness and tolerance for occasional deadline misses.

@@

\section{Trace-cmd}\label{sec:trace-cmd}
\texttt{Trace-cmd}~\cite{Tracecmd} is the front-end tool of the \texttt{ftrace} tool that is built inside the Linux kernel and used for tracing the Linux kernel. \texttt{Trace-cmd} can record detailed reports of kernel events such as interrupts, scheduler decisions, function calls, and custom events in real time. It is easier with this tool to enable and disable tracing, set filters, and view results. \texttt{Trace-cmd} helped debugging and identifying the reasons for latency in Salamander 4 to continuously improve its real-time performance and determinism. A very practical aspect of \texttt{trace-cmd} is that the guest can be traced from the host. For this purpose, the guest needs to be configured up with \texttt{vsocks}, which are a virtual socket that enable direct communication between the host and guest. To enable vsockets and tracing, the guest kernel requires the following configurations, according to Linux developer and maintainer Steve Rostedt~\cite{rostedtTracecmdHostGuest}.

@@

\section{Approach}\label{sec:approach}
The goal is to virtualize the Salamander 4 OS and bring its real-time performance closer to that of the hardware version and guarantee deterministic and reliable behavior. For that purpose, the first step is to apply the PREEMPT-RT patch to the host system and subsequently tune the real-time performance through BIOS, kernel, host, guest, and QEMU/KVM configurations. These modifications are added sequentially and tested together. First, the BIOS is configured, followed by the kernel. The BIOS settings are not reverted when moving to the kernel configurations. Next, the host is configured, but the BIOS and kernel settings remain unchanged. The guest OS is already based on Xenomai and therefore, no additional modifications are necessary for the guest OS itself. Finally, QEMU/KVM settings are applied. This means that the configurations are not isolated but are applied and then tested as a whole. The key metric here is the latency of the system, and the real-time performance is evaluated using the \texttt{latency} tool of the Xenomai suite after each configuration step. At the end, the results are demonstrated altogether and a holistic comparison and discussion are carried out. In addition, the improvement in real-time performance and determinism is demonstrated using a robotic application, where the modified virtualization is compared with the hardware version.

>>>>

%Formatieren des Quellcodeverzeichnisses
\makeatletter
% Setzen der Bezeichnungen für das Quellcodeverzeichnis/Abkürzungsverzeichnis in Abhängigkeit von der eingestellten Sprache
\providecommand\listacroname{}
\@ifclasswith{twbook}{english}
{%
    \renewcommand\lstlistingname{Code}
    \renewcommand\lstlistlistingname{List of Codes}
    \renewcommand\listacroname{List of Abbreviations}
}{%
    \renewcommand\lstlistingname{Quellcode}
    \renewcommand\lstlistlistingname{Quellcodeverzeichnis}
    \renewcommand\listacroname{Abkürzungsverzeichnis}
}
% Wenn die Option listof=entryprefix gewählt wurde, Definition des Entyprefixes für das Quellcodeverzeichnis. Definition des Macros listoflolentryname analog zu listoflofentryname und listoflotentryname der KOMA-Klasse
\@ifclasswith{scrbook}{listof=entryprefix}
{%
    \newcommand\listoflolentryname\lstlistingname
}{%
}
\makeatother
\newcommand{\listofcode}{\phantomsection\lstlistoflistings}

@@

\bigskip \noindent Insgesamt bietet diese Arbeit eine umfassende Methodik, um ein virtualisiertes Gastsystem in einem Hostsystem mit deterministischem Verhalten echtzeitfähig zu machen.}
\schlagworte{Virtualisierung, Echtzeitsysteme, Latenzreduktion, Robotersteuerung}

@@

\chapter{Introduction}\label{cha:introduction}
In today's industrial production and automation, robot systems are well established and of crucial importance. Robots must react to their environment and perform time-critical tasks within strict time constraints. Delays or errors can have catastrophic consequences in some cases. Modern operating systems nowadays are categorized into two types:  General-Purpose Operating System (GPOS) and Real-Time Operating System (RTOS). The key difference is whether the system is time-critical or not~\cite{canbazPerformanceAnalysisRealtime2022}. In a traditional GPOS, such as Windows or Linux-based distributions, a high-priority task cannot interrupt a kernel operation, as illustrated in Figure~\ref{fig:kernel_generic}. This makes them not suitable for real-time requirements, as they cannot guarantee deterministic execution times.

@@

\noindent In these real-time operating systems, task scheduling is based on so-called priority-based preemptive scheduling~\cite{buttazzoHardRealtimeComputing2024}. Each task in a software application is assigned a priority. A higher priority means that a faster response is required. Preemptive task scheduling ensures a very fast response. Preemptive means that the scheduler can stop a currently running task at any point if it recognizes that another task needs to be executed immediately. The basic rule on which priority-based preemptive scheduling is based is that the task with the highest priority that is ready to run is always the task that must be executed. So, if both a task with a lower priority and a task with a higher priority are ready to run, the scheduler ensures that the task with the higher priority runs first. The lower-priority task is only executed once the higher priority task has been processed.

@@

\begin{itemize}
	\item This work was written at SIGMATEK GmbH \& Co KG, a company that researches, develops, produces, and distributes automation technology for industrial machinery and plant engineering. SIGMATEK uses its own custom Linux-based operating system, to be run on their self-manufactured CPUs. This operating system will be referred to as ``Salamander 4'' in this work. The details of Salamander 4 are explained in Section~\ref{sec:guest_operating_system}.
	\item Salamander 4 is virtualized through Quick Emulator (QEMU), an open-source emulator and virtualizer that allows running different operating systems on a computer. QEMU is described in Section~\ref{sec:qemu}.
	\item Salamander 4 was created with Yocto, an open-source project that helps develop custom Linux-based systems. Yocto is discussed in Section~\ref{sec:yocto}.
	\item Salamander 4 employs hard real-time with Xenomai 3, a real-time framework for Linux that enables deterministic and low-latency performance. Xenomai 3 is detailed in Section~\ref{sec:xenomai}.
	\item The goal is to virtualize Salamander 4 and approach the performance of bare metal (hardware) through real-time performance tunings, focusing on latency requirements set by SIGMATEK. These modifications are the subject of Section~\ref{sec:real-time_tunings}.
	\item The \texttt{latency} tool of the Xenomai suite to measure the scheduling latency of the real-time OS was suggested to be used by SIGMATEK. The goal was set to achieve latency values below 50 microseconds in the duration of the measurement.
	\item SIGMATEK uses the VARAN bus for communication between machines and systems, a real-time Ethernet network that was specially developed for industrial automation. Its operating principle is outlined in Section~\ref{sec:varan}.
	\item For the purpose of replacing the physical CPU, SIGMATEK provided a PCI Insert Card module PCV 522, that was plugged into the PC, as explained in Subsection~\ref{subsec:setup_experiment_virtualized}. This component could then be used to communicate with the Input/Output (I/O) peripherals, just like the physical CPU could when the I/O modules were attached to it. Subsequently, the PCV 522 module of SIGMATEK was connected with the Pulse Width Module PW 022 of SIGMATEK over the VARAN bus and the Varan Connection Module VI 021 of SIGMATEK.
	\item The robotic application from Section~\ref{sec:robotic_application} was written in Lasal Class 2, an object-oriented programming tool of SIGMATEK and part of SIGMATEK's Engineering Platform LASAL, with client-server technology and graphic representation.
\end{itemize}

@@

\bigskip \noindent \citeauthor{perneelRealtimeCapabilitiesStandard2015}~\cite{perneelRealtimeCapabilitiesStandard2015} mention in their work that Linux was initially developed as a general-purpose operating system without real-time applications in mind. They talk about the evolution of real-time behavior in the standard Linux kernel and state that it has recently gained popularity among the real-time community, largely due to its open-source nature and stability. The authors explain how to transform Linux into a real-time operating system by implementing the PREEMPT-RT patch. The paper also emphasizes some kernel configurations that ensure that the system's real-time behavior is preserved during runtime. Even though these modifications to the kernel enable soft real-time performance, the authors underline the fundamental rule in real-time software, that latency improvements have a negative impact on throughput and kernel performance. This is due to the added overhead of the \texttt{CONFIG\_PREEMPT\_RT} option. They concluded with their Linux build and a testing application that Linux can be a viable option for real-time applications, if it is correctly configured. \citeauthor{reghenzaniRealTimeLinuxKernel2020}~\cite{reghenzaniRealTimeLinuxKernel2020} provide an extensive overview of the real-time Linux kernel research, specifically on the PREEMPT RT patch. \citeauthor{adamRealTimePerformanceResponse2021}~\cite{adamRealTimePerformanceResponse2021} applied the PREEMPT\_RT patch on both Raspberry Pi 3 and BeagleBone Black and achieved significantly lower latencies compared to standard Linux kernels.

@@

\bigskip \noindent In the course of writing this thesis, the white paper "Real-Time Performance Tuning Best Practice Guidelines for KVM-Based Virtual Machines"~\cite{RealTimePerformanceTuning2022} by Intel has been instrumental. Given that the working computer used for this work is equipped with an Intel processor, the guidelines presented in this document were particularly relevant and valuable for optimizing real-time performance in KVM-based virtual machines. Section~\ref{sec:real-time_tunings} contains an extensive tuning process, which involves configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS itself. The document concludes that the described tuning methods effectively improve real-time performance for KVM-based VMs, even under heavy workloads.

@@

\bigskip \noindent \citeauthor{mckenneyRealTimeVs2008}~\cite{mckenneyRealTimeVs2008} addresses the overheads associated with real-time Linux when using KVM and QEMU. The author lists several sources of overhead, including memory locking to avoid page-fault latencies, increased overhead of locking and interrupts due to more-aggressive preemption, threaded interrupts that permit long-running interrupt handlers to be preempted by high-priority real-time processes, real-time task scheduling that requires global scheduling, high-resolution timers for tens-of-microseconds accuracy and precision, and preemptible RCU which has slightly higher read-side overhead than classic RCU. This information helped understand the trade-offs involved in real-time Linux systems.

\bigskip \noindent \citeauthor{deoliveiraDemystifyingRealTimeLinux2020}~\cite{deoliveiraDemystifyingRealTimeLinux2020} critique the traditional \texttt{cyclictest} tool used for measuring scheduling latency, highlighting its limitations due to its black-box approach and lack of theoretical grounding. This contributed to the decision not to use this tool to measure the scheduling latency in this work. Instead, the \texttt{latency} tool was used, as was determined in Section~\ref{sec:application_context}.

@@

\bigskip \noindent Virtualization promises to overcome these challenges~\cite{bhardwajVirtualizationCloudComputing2021}~\cite{queirozContainerbasedVirtualizationRealtime2024}~\cite{cinqueEvaluatingVirtualizationFog2023}. However, virtualization also has its own challenges and pitfalls. The additional layers of abstraction can lead to increased overhead and latency~\cite{casiniLatencyAnalysisVirtualization2021}~\cite{zhangEvaluatingOptimizingVirtualization2010}. This is not desired and can be particularly problematic for real-time systems that require strict timing constraints. Making sure that tasks run predictably and on time in a virtualized environment can be challenging. The hypervisor and other virtual components can cause variations in how long tasks take to execute~\cite{garcia-vallsChallengesRealtimeVirtualization2014}. When there are multiple virtual machines (VMs) sharing the same physical resources, the VMs can affect each other's performance and consequently violate real-time constraints~\cite{queirozContainerbasedVirtualizationRealtime2024}.

@@

\bigskip \noindent In this master's thesis, the task is the virtualization of a real-time operating system to control a robot, with a focus on compliance with real-time determinism. Chapter~\ref{cha:methodology} explains in detail the procedure, how this was tackled. As already established in Section~\ref{sec:application_context}, Salamander 4 is built with Yocto, employs hard real-time with Xenomai 3, and is virtualized through QEMU/KVM. The latency values are measured with the \texttt{latency} program of the Xenomai tool suite. The initial situation is described in Section~\ref{sec:starting_situation}. In a nutshell, there is a significant initial gap in latency between the virtualized Salamander 4 and the Salamander 4 on bare metal. For this reason, an extensive tuning process is carried out to achieve real-time performance and determinism. These modifications involve configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 guest OS itself. The individual configurations are discussed in detail in Section~\ref{sec:real-time_tunings} and the modifications are justified with a clear explanation.

\section{Objective}
The goal is to bring the latency of the virtualized Salamander 4 closer to that of the hardware (bare metal), thereby providing a comprehensive blueprint for making a virtualized guest system in a host system real-time capable with deterministic behavior. Section~\ref{sec:starting_situation} outlines the initial situation of this master's thesis and serves as the foundation for the real-time performance tunings discussed in Section~\ref{sec:real-time_tunings}. The objective is to achieve latency values below 50 microseconds during the measurement period through real-time tunings, with no outliers whatsoever. To determine and validate the performance of the tunings, the \texttt{latency} tool is used, as stated in Section~\ref{sec:application_context}. On top of that, the aim is to use a robotic application in a practical scenario to demonstrate and validate that the tunings significantly reduce latency and improve determinism compared to the unmodified virtualization.

@@

This chapter describes in detail all the theoretical concepts and practical methods that contributed to achieving the objectives of this master's thesis.

@@

\subsection{Structure}
\noindent Salamander 4 is the proprietary operating system of SIGMATEK. It is based on Linux version 5.15.94 and integrates Xenomai 3.2, a real-time development environment. Salamander 4 is a 64-bit system, which refers to the x86\_64 architecture. The real-time behaviour is achieved through the use of Symmetric Multi-Processing (SMP) and Preemptive Scheduling (PREEMPT). In addition, it uses IRQPIPE to process interrupts in a way that meets the real-time requirements of the system. The output of the command~\texttt{uname -a} can be observed in Code~\ref{output:uname_a_virt}.

\vspace{2em}
\begin{lstlisting}[name={Guest Operating System Information},label={output:uname_a_virt}]
root@sigmatek-core2:~# uname -a
Linux sigmatek-core2 5.15.94 #1 SMP PREEMPT IRQPIPE Tue Feb 14 18:18:05 UTC 2023 x86_64 GNU/Linux
\end{lstlisting}

@@

\begin{itemize}
	\item \textbf{Operating system}: The operating system in a LASAL CPU manages the hardware and software resources of the system. It is provided in a completely PC-compatible manner, working with a standard PC BIOS.
	\item \textbf{Loader}: The loader is a part of the operating system that is responsible for loading programs from executables into memory, preparing them for execution and then running them.
	\item \textbf{Hardware classes}: Hardware classes in LASAL represent the different types of hardware components that can be controlled by the LASAL CPU. They help organize and manage the hardware components in a modular manner. The graphical hardware editor in LASAL allows for a accurate simulation of the actual hardware.
	\item \textbf{Applications}: Applications are developed using LASAL CLASS 2~\cite{EngineeringToolLASAL}, a solution for automation tasks that supports object-oriented programming and design in compliance with IEC 61131-3.
\end{itemize}

@@

Quick Emulator (QEMU)~\cite{QEMU} is an open-source emulator and virtualizer that allows running operating systems and programs for one machine on a different machine. It can emulate different hardware architectures and is used in this thesis to emulate Salamander 4. When QEMU is used together with Kernel-based Virtual Machine (KVM), it can provide near-native performance by running code directly on the host CPU. The QEMU script used to emulate Salamander 4 and the meaning of the options are described in detail in Subsection \ref{subsec:salamander4-virtualization}.

@@

\vspace{2em}
\begin{lstlisting}[name={Contents of QEMU Folder for Salamander 4},label={lst:ls}]
    sigma_ibo@localhost:~/build/tmp/deploy/qemu/salamander-image$ ls -1
    bzImage
    drive-c
    ovmf.code.qcow2
    qemu_def.sh
    salamander-image-sigmatek-core2.ext4
    stek-drive-c-image-sigmatek-core2.tar.gz
    vmlinux
    \end{lstlisting}

@@

\begin{itemize}
	\item \textbf{bzImage}: Compressed Linux kernel image that is loaded by QEMU at system start. The short form ``bz`` stands for big-zipped.
	\item \textbf{drive-c}: Directory serving as C drive for QEMU, created and filled by \texttt{qemu\_def.sh} script.
	\item \textbf{ovmf.code.qcow2}: Firmware file for QEMU that enables UEFI boot process. ~\texttt{OVMF} stands for Open Virtual Machine Firmware,~\texttt{qcow2} is a format for disk image files used by QEMU, it stands for ``QEMU Copy On Write version 2''.
	\item \textbf{qemu\_def.sh}: Shell script that starts QEMU with required parameters to boot the Salamamder 4 OS. It is described later in more detail.
	\item \textbf{salamander-image-sigmatek-core2.ext4}: Disk image of the Salamander 4 OS. It uses the~\texttt{ext4} file system and serves as the root file system in the QEMU virtual machine, acting as the virtual hard drive.
	\item \textbf{stek-drive-c-image-sigmatek-core2.tar.gz}: Compressed tarball containing a pre-configured environment for the Salamander 4 OS. It is unpacked and sets up the~\texttt{drive-c/} directory with system and log files in the~\texttt{qemu\_def.sh} script.
	\item \textbf{vmlinux}: Uncompressed Linux kernel image, typically used for debugging.
\end{itemize}

\noindent The initial QEMU script after the custom Yocto build and the starting point for this work is shown in Code~\ref{script:qemu_def}. This script is used to start QEMU with required parameters to boot the Salamander 4 OS. It will be adjusted in Section~\ref{sec:real-time_tunings} in order to accompany real-time performance tunings.

\vspace{2em}
\begin{lstlisting}[name={QEMU Script for starting Salamander 4 Virtualization},label={script:qemu_def}]
	#!/bin/sh

@@

\noindent  The second approach, called Mercury and shown in Figure~\ref{fig:mercury}, relies on the real-time capabilities already present in the native Linux kernel. Often, applications require the PREEMPT-RT patch~\cite{RealtimePreempt_rt_versionsWiki} to augment the mainline kernel's real-time responsiveness and minimize jitter, but this is not mandatory and depends on the application's specific requirements for responsiveness and tolerance for occasional deadline misses.

@@

\section{Trace-cmd}\label{sec:trace-cmd}
\texttt{Trace-cmd}~\cite{Tracecmd} is the front-end tool of the \texttt{ftrace}~\cite{FtraceFunctionTracer} tool that is built inside the Linux kernel and used for tracing the Linux kernel. \texttt{Trace-cmd} can record detailed reports of kernel events such as interrupts, scheduler decisions, function calls, and custom events in real time. It is easier with this tool to enable and disable tracing, set filters, and view results. \texttt{Trace-cmd} helped debugging and identifying the reasons for latency in Salamander 4 to continuously improve its real-time performance and determinism. A very practical aspect of \texttt{trace-cmd} is that the guest can be traced from the host. For this purpose, the guest needs to be configured with \texttt{vsocks}, which are a virtual socket that enable direct communication between the host and guest. To enable vsockets and tracing, the guest kernel requires the following configurations, according to Linux developer and maintainer Steve Rostedt~\cite{rostedtTracecmdHostGuest}.

@@

\section{Approach}\label{sec:approach}
The goal is to virtualize the Salamander 4 OS and bring its real-time performance closer to that of the bare metal version and guarantee deterministic and reliable behavior. For that purpose, the first step is to apply the PREEMPT-RT patch to the host system and subsequently tune the real-time performance through BIOS, kernel, host, QEMU/KVM, and guest configurations. These modifications are added sequentially and tested together. First, the BIOS is configured, followed by the kernel. The BIOS settings are not reverted when moving to the kernel configurations. Next, the host is configured, but the BIOS and kernel settings remain unchanged. The guest OS is already based on Xenomai and therefore, no additional modifications are necessary for the guest OS itself. Finally, QEMU/KVM settings are applied. This means that the configurations are not isolated but are applied and then tested as a whole. The key metric here is the latency of the system, and the real-time performance is evaluated using the \texttt{latency} tool of the Xenomai suite after each configuration step. At the end, the results are demonstrated altogether and a holistic comparison and discussion are carried out. In addition, the improvement in real-time performance and determinism is demonstrated using a robotic application, where the tuned virtualization is compared with the untuned and the bare metal version.
